---
title: "Asset indices with MCA"
author: "Peter S. Larson"
date: "2021/12/29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Asset indices with Multiple Correspondence Analysis

There are many methods of creating SES indices. Traditionally, people have used PCA (See Filmer, Pritchett, 2001). PCA, however, assumes normally distributed continuous data. Asset surveys however are almost invariably categorical. In practice, it might not make much of a difference, but since PCA is inappropriate and other methods exist, I recommend trying something else. 

Multple Correspondence Analysis (MCA) is intended for categorical data. 

## Get some data

First, let's get some data. We'll be creating fake data which won't have all the properties of real data, but is created to at least resemble what a real data set would look like. 

We'll add some libraries that we'll need, create the data and assess missingness. In this example, our data set won't have any missing data. If you would like to see an example with missing data, please wait until I can write one up. 

```{r get_ses_data, echo=FALSE}

#### Get require libraries ####

require(dplyr)
require(readr)
require(janitor)
require(VIM)
require(ggplot2)
require(MASS)
require(gtools)

#### Set data path and load in data ####

source("ses_make_some_fake_data.R")

## Get the data set but only with the asset columns (no ID)
ses_data_only_asset_columns <- ses_data %>%
  dplyr::select(-ID) %>%
  mutate_at(names(ses_data %>%
                    dplyr::select(-ID)), funs(factor(.))) ## Convert everything to factors


#### Check for missingness ####

aggr(ses_data_only_asset_columns, numbers = TRUE, prop = c(TRUE, FALSE))

## No missing data, proceed
```

## Running the MCA

Here will will be using the mca() function from the FactoMineR package. We will run the MCA, create a useful plot, and then add the first component to our data set.


```{r mca, echo=FALSE}
require(FactoMineR)

# apply mca
mca = mca(ses_data_only_asset_columns, nf = 5)

### Plot the results of the MCA to see what it all looks like 

# data frame for ggplot

## Set number of categories for each variable, though in this case there are only two.
cats = apply(ses_data_only_asset_columns , 2, function(x) nlevels(as.factor(x)))

### Put this all into a data frame
mca2_vars_df = data.frame(mca$cs, Variable = rep(names(cats), cats))

# data frame with observation coordinates
mca1_obs_df = data.frame(mca$ind$coord)


# plot to see if what we are doing matches our intuition. Does it?
ggplot(data = mca2_vars_df, 
       aes(x = X1, y = X2, label = rownames(mca2_vars_df))) +
  geom_hline(yintercept = 0, colour = "gray70") +
  geom_vline(xintercept = 0, colour = "gray70") +
  geom_text(aes(colour = Variable)) +
  ggtitle("Loadings for the first two MCA components of Ecodess SES by variable")



#### Add first component back to original data file to get our asset index
ses_data$ses_mca <- mca$rs[,1]

#### Cut into categories and rename ####
ses_data$ses_categories <- quantcut(ses_data$ses_mca, 5)
levels(ses_data$ses_categories) <- c("5 (Wealthiest)", "4","3","2", "1 (Poorest)")

## Summarize
summary(ses_data)
```

## Two component version

If want to use two components instead of one, you can just add them together. 

```{r two_components}

#### Add first component back to original data file to get our asset index
ses_data$ses_mca_1 <- mca$rs[, 1]

#### Add second component
ses_data$ses_mca_2 <- mca$rs[, 2]

#### Sum both of these ####
ses_data$ses_mca <- ses_data$ses_mca_1 + ses_data$ses_mca_2

#### Cut into categories and rename ####
ses_data$ses_categories <- quantcut(ses_data$ses_mca, 5)
levels(ses_data$ses_categories) <-
  c("5 (Wealthiest)", "4", "3", "2", "1 (Poorest)")

## Summarize
summary(ses_data)

### Plot summed components with component 1 or 2
ggplot(data = ses_data,
       aes(x = ses_mca, y = ses_mca_1)) +
  geom_point() +
  ggtitle("MCA two component score with component 1") +
  xlab("Two component SES composite") +
  ylab("MCA component 1")


### Plot summed components with component 1 or 2
ggplot(data = ses_data,
       aes(x = ses_mca, y = ses_mca_2)) +
  geom_point() +
  ggtitle("MCA two component score with component 2") +
  xlab("Two component SES composite") +
  ylab("MCA component s")



```
